% Altimeter Pipeline — Validation Results Section
% Drop into your paper's results section.

\subsection{Pipeline validation}

\subsubsection{Cross-site QC performance}

The pipeline was validated across five representative deployments spanning
all three sites, both instrument types (AA400 altimeter and EA400
echosounder), and three sampling modes (continuous, burst-300, burst-60).
Table~\ref{tab:alt_validation} summarizes the QC performance.

\begin{table}[h]
\centering
\caption{Pipeline validation across representative deployments.
IQR = median within-burst interquartile range.}
\label{tab:alt_validation}
\begin{tabular}{lcccccc}
\toprule
Deployment & Instrument & Bursts & Rejected & BL range & IQR & Valid\% \\
\midrule
SIO24 (6m, continuous) & AA400 & 257 & 34 & 71~mm & 4.6~mm & 97\% \\
SIO24 (6m, burst-300) & AA400 & 1531 & 6 & 57~mm & 5.1~mm & 97\% \\
TP24 (10m, echosounder) & EA400 & 1719 & 70 & 94~mm & 7.2~mm & 94\% \\
TP25 (10m, burst-60) & AA400 & 6883 & 275 & 128~mm & 3.8~mm & 97\% \\
SOL25 (7m, tilted) & EA400 & 2600 & 490 & 99~mm & 7.3~mm & 88\% \\
\bottomrule
\end{tabular}
\end{table}

The AA400 altimeters consistently exhibited lower within-burst IQR
(\SIrange{3.8}{5.1}{\milli\meter}) compared to the EA400 echosounders
(\SIrange{7.2}{7.3}{\milli\meter}), reflecting the simpler measurement
geometry and more focused acoustic beam of the dedicated altimeter.
The percent of valid samples within accepted bursts exceeded 88\%
across all deployments. The Solana Beach deployment demonstrated that
the tilt-deviation mask correctly preserves data from a tilted
instrument (baseline tilt \SI{13.2}{\degree}) while rejecting periods
of transient disturbance.

\subsubsection{Firmware error recovery}

The Torrey Pines 5m altimeter (Nov 2023 -- Feb 2024) was affected by
a firmware bug that overwrote valid altitude data with date-string
error messages after memory filled ($\sim$113~days). The two-level
despiking approach recovered the underlying morphological signal:
the per-ping phase-space despike removed individual corrupted samples
within bursts, and the iterative burst-level despike (5~passes)
removed 1472~additional corrupted bursts. The resulting time series
clearly shows the \SI{700}{\milli\meter} accretion event caused by
the December 28, 2023 storm, consistent with the subaerial beach
erosion observed in concurrent surveys.

\subsubsection{Independent validation against beach surveys}

Echosounder bed level measurements were independently validated
against beach morphology surveys conducted by the Scripps Coastal
Processes Group using jetski-mounted RTK GPS systems. Survey profiles
were interpolated to the instrument cross-shore position at each
survey date.

For single-deployment validation at Solana Beach (March--June 2025),
the RMSE of the bed level change between consecutive survey pairs was
\SI{22}{\milli\meter} with a bias of \SI{-4}{\milli\meter}. This is
comparable to the spatial variability inherent in the survey comparison
itself (survey-to-survey $\sigma = \SI{21}{\milli\meter}$), indicating
that the instrument accuracy is at least as good as can be resolved by
the available ground truth.

For multi-deployment validation spanning the full record at each site,
the cross-deployment baseline anchoring introduced additional
uncertainty. Torrey Pines 10m (survey-anchored, 3~pairs) showed
RMSE = \SI{61}{\milli\meter}. The Solana Beach multi-deployment record
showed larger RMSE (\SI{92}{\milli\meter}), attributable to physical
relocation of the echosounder pipe during maintenance (re-jetting),
which changed the spatial point of measurement between deployments.

\subsubsection{L4 merged products}

Full-site L4 merged products were constructed for SIO~pier at
\SI{6}{\meter} and Torrey Pines at \SI{5}{\meter}, \SI{10}{\meter},
and \SI{15}{\meter}. PUV L2 spectral segments were matched to
altimeter bursts by nearest-neighbor within $\pm$\SI{5}{\minute}
(median offset \SI{2.5}{\minute}). Continuous hourly $H_s$ from the
CDIP MOP spectral wave model was interpolated to burst timestamps to
fill PUV deployment gaps, providing 100\% wave height coverage over
the full altimeter record at all sites.

% Table: L4 coverage
\begin{table}[h]
\centering
\caption{L4 merged product coverage.}
\label{tab:l4_coverage}
\begin{tabular}{lcrrr}
\toprule
Site & Depth & Altimeter bursts & PUV matched (\%) & MOP gap-fill \\
\midrule
SIO Pier & 6~m & 32{,}267 & 8{,}190 (25\%) & 25{,}424 \\
Torrey Pines & 5~m & 32{,}486 & 5{,}560 (17\%) & 27{,}998 \\
Torrey Pines & 10~m & 25{,}696 & 9{,}031 (35\%) & 18{,}175 \\
Torrey Pines & 15~m & 24{,}092 & 8{,}352 (35\%) & 16{,}415 \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Sediment mobilization and transport thresholds}

\subsubsection{Depth-dependent mobilization}

Using the \citet{soulsby1997} critical Shields parameter for
$D_{50} = \SI{0.25}{\milli\meter}$ (fine sand, $D_* = 6.0$,
$\theta_{cr} = 0.043$, $\tau_{cr} = \SI{0.17}{\pascal}$), the
fraction of burst-averaged segments exceeding the mobilization
threshold decreases monotonically with depth: 100\% at \SI{5}{\meter},
95\% at \SI{6}{\meter}, 87\% at \SI{10}{\meter}, and 69\% at
\SI{15}{\meter} (Table~\ref{tab:thresholds}). At the shallowest
sites (5--6~m), the bed is in near-continuous transport; at
\SI{15}{\meter}, approximately one-third of observations are below
the theoretical mobilization threshold.

\begin{table}[h]
\centering
\caption{Sediment mobilization and transport characteristics by depth.
$\theta_{med}$: median Shields parameter during PUV-matched bursts.
Transport efficiency: mean $|d\Delta z/dt|$ divided by mean $\tau_b$.}
\label{tab:thresholds}
\begin{tabular}{lccccc}
\toprule
Site & Depth & $\theta_{med}$ & Mobilized & Mean $|d\Delta z/dt|$ & Efficiency \\
 & (m) & & (\%) & (mm/hr) & (mm/hr/Pa) \\
\midrule
SIO Pier & 6 & 0.105 & 95 & 5.8 & 11.5 \\
Torrey Pines & 5 & 0.240 & 100 & 26.9 & 26.1 \\
Torrey Pines & 10 & 0.091 & 87 & 9.4 & 20.6 \\
Torrey Pines & 15 & 0.057 & 69 & 10.2 & 34.9 \\
\bottomrule
\end{tabular}
\end{table}

\subsubsection{Transport rate and excess shear stress}

The relationship between bed level change rate and excess Shields
parameter ($\theta - \theta_{cr}$) was examined by fitting power
laws of the form $|d\Delta z/dt| = a\,(\theta - \theta_{cr})^n$ to
the binned data at each depth. Classical bedload transport formulas
predict $n \approx 1.5$ \citep[Meyer-Peter M\"uller,][]{soulsby1997};
however, the observed exponents were dramatically lower:

\begin{itemize}
  \item Torrey Pines 5~m: $n = 0.16$ ($a = 33.9$~mm/hr)
  \item SIO Pier 6~m: $n = -0.04$ ($a = 5.0$~mm/hr)
  \item Torrey Pines 10~m: $n = -0.01$ ($a = 9.0$~mm/hr)
  \item Torrey Pines 15~m: $n = -0.02$ ($a = 9.1$~mm/hr)
\end{itemize}

The near-zero exponents indicate that the bed level change rate is
approximately \textbf{independent of excess shear stress} once the
threshold is exceeded. This suggests a transport-capacity-saturated
regime in which the rate of morphological change is limited not by
the available forcing but by sediment supply, bedform dynamics, or
gravitational restoring forces. The \SI{5}{\meter} site shows the
strongest stress dependence ($n = 0.16$), consistent with its
proximity to the surf zone where wave nonlinearity and turbulence
generate additional transport mechanisms beyond those captured by
the Shields parameterization.

\subsubsection{Transport efficiency}

The transport efficiency, defined as mean $|d\Delta z/dt|$ per unit
mean $\tau_b$, increases with depth: \SI{11.5}{\milli\meter\per\hour
\per\pascal} at \SI{6}{\meter} to \SI{34.9}{\milli\meter\per\hour
\per\pascal} at \SI{15}{\meter}. This depth dependence is consistent
with finer sediment at greater depths (lower threshold, more easily
mobilized per unit stress) and is opposite to the total bed change
rate, which decreases with depth because less wave energy reaches the
bed. The \SI{5}{\meter} site is an outlier with both high total
transport rate (\SI{26.9}{\milli\meter\per\hour}) and moderately high
efficiency (\SI{26.1}{\milli\meter\per\hour\per\pascal}), reflecting
the energetic surf-zone-adjacent conditions.

\subsubsection{Accretion--erosion asymmetry}

Separating the signed bed level change rate into accretion
($d\Delta z/dt > 0$) and erosion ($d\Delta z/dt < 0$) components
reveals a depth-dependent asymmetry. At the \SI{5}{\meter} site,
erosion rates at high $\tau_b$ significantly exceed accretion rates,
indicating that strong wave events preferentially remove sand from
this depth. At deeper sites (\SI{10}{\meter}, \SI{15}{\meter}),
accretion and erosion magnitudes are approximately symmetric,
consistent with a more passive morphological response to oscillatory
forcing.

\subsubsection{Velocity moment correlations}

The relationship between bed level change and wave velocity moments
\citep{hoefel2003} was assessed using the PUV-derived velocity
skewness and the third-order moment $\langle u|u|^2 \rangle$. At
Torrey Pines \SI{5}{\meter}, the $\langle u|u|^2 \rangle$ moment
shows a clear correlation with signed $d\Delta z/dt$: positive
velocity moments (onshore-skewed wave forcing) correspond to positive
bed level change (accretion), and negative moments correspond to
erosion. This is consistent with the \citet{hoefel2003} energetics
formulation, in which the net cross-shore sediment flux depends on
the third moment of the near-bed velocity.

At deeper sites (SIO \SI{6}{\meter}, Torrey Pines \SI{10}{\meter}
and \SI{15}{\meter}), the velocity moment signal is weaker and
noisier, indicating that wave nonlinearity is less effective at
driving net transport at these depths. This depth dependence is
expected: wave skewness and asymmetry decrease with increasing
depth as waves become more sinusoidal farther from the surf zone.

\subsection{Equilibrium model critique}

The Dean/Yates/Ludka equilibrium beach change framework was tested
against the depth-resolved L4 observations. Three formulations were
evaluated:

\begin{enumerate}
  \item \textbf{Dean $\Omega$ disequilibrium}: $d\Delta z/dt$ versus
    $\Omega - \Omega_{eq}$, where $\Omega = H_s / (w_s T_p)$ and
    $\Omega_{eq}$ is the deployment median.
  \item \textbf{Wave energy proxy}: $d\Delta z/dt$ versus $H_s^2$.
  \item \textbf{State dependence}: $d\Delta z/dt$ versus current bed
    level $\Delta z$ (the Yates/Ludka mechanism: above equilibrium
    $\rightarrow$ erosion, below $\rightarrow$ accretion).
\end{enumerate}

All three formulations failed at all four depths, with $R^2 < 0.01$
in every case (Table~\ref{tab:equilibrium}).

\begin{table}[h]
\centering
\caption{Equilibrium model test results. All $R^2$ values $< 0.01$.}
\label{tab:equilibrium}
\begin{tabular}{lcccc}
\toprule
Site & Depth & $\Omega$ diseq.\ $R^2$ & $H_s^2$ $R^2$ & State $R^2$ \\
\midrule
SIO Pier & 6~m & 0.006 & 0.010 & 0.000 \\
Torrey Pines & 5~m & 0.000 & 0.000 & 0.008 \\
Torrey Pines & 10~m & 0.003 & 0.000 & 0.001 \\
Torrey Pines & 15~m & 0.005 & 0.005 & 0.000 \\
\bottomrule
\end{tabular}
\end{table}

The equilibrium concept---that $d\Delta z/dt$ is proportional to the
deviation from a wave-determined equilibrium state---has essentially
zero predictive power for bed level change rate at \SIrange{5}{15}{\meter}
depth. This is consistent with the capacity-saturation finding
($n \approx 0$): once the mobilization threshold is exceeded, the
transport rate does not increase with forcing magnitude, so the
disequilibrium metric carries no information about the rate of
morphological change.

\subsection{Storm event response}

The December 28, 2023 storm (peak $H_s = \SI{3.2}{\meter}$ at
\SI{10}{\meter}$ depth) was analyzed as a case study of depth-resolved
storm morphological response across the Torrey Pines cross-shore
transect (Table~\ref{tab:storm}).

\begin{table}[h]
\centering
\caption{December 28, 2023 storm response at Torrey Pines MOP586.
Net $\Delta z$ = post-storm minus pre-storm median bed level.}
\label{tab:storm}
\begin{tabular}{lcccc}
\toprule
Depth & Pre-storm & Post-storm & Net $\Delta z$ & Peak $|d\Delta z/dt|$ \\
(m) & (mm) & (mm) & (mm) & (mm/day) \\
\midrule
5 & $-160$ & $+805$ & $\mathbf{+965}$ & 1500 \\
7 & $-169$ & $+155$ & $+324$ & 600 \\
10 & $-40$ & $-3$ & $+38$ & 120 \\
15 & $-23$ & $-23$ & $-1$ & 30 \\
\bottomrule
\end{tabular}
\end{table}

The storm deposited nearly \SI{1}{\meter} of sand at \SI{5}{\meter}
depth (the storm bar), with the deposit attenuating exponentially
across the profile: one-third as much at \SI{7}{\meter},
\SI{38}{\milli\meter} at \SI{10}{\meter}, and zero net change at
\SI{15}{\meter}. The \SI{15}{\meter} site acted as the effective
depth of closure for this storm magnitude. Peak bed change rates at
\SI{5}{\meter} reached \SI{1500}{\milli\meter\per\day}
(\SI{1.5}{\milli\meter\per\minute}) during the storm peak.
